---
title: "Group 8 Final Code Demonstration"
author: "STATS 101A Group 8"
date: "2018/3/22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Data cleaning (NOTICE: A different approach will be used for the testing dataset)


```{r data_cleaning}
setwd("~/Downloads/STATS 101A")
happy<-read.table("Happiness.txt",header=T)
library(dplyr)

#First, eliminate all the cases where Happy Value is missing
happy1 <- happy %>% filter(Happy!=8&Happy!=9)

#Change all the undefined values into NA
happy1[(happy1$Household==8 | happy1$Household==9 | 
          happy1$Household==0),]$Household<-NA
happy1[(happy1$Health==8 | happy1$Health==9 | happy1$Health==0),]$Health<-NA
happy1[(happy1$OwnHome==8 | happy1$OwnHome==9 | happy1$OwnHome==0),]$OwnHome<-NA
happy1[(happy1$Instagram==8 | happy1$Instagram==9 | happy1$Instagram==0),]$Instagram<-NA

happy1[(happy1$Marital==9),]$Marital<-NA
happy1[(happy1$Age==98|happy1$Age==99),]$Age<-NA
happy1[(happy1$Children==9),]$Children<-NA
happy1[(happy1$Education==98 | happy1$Education==99 | happy1$Education==97),]$Education<-NA
happy1[(happy1$JobSat==8 | happy1$JobSat==9 | happy1$JobSat==0),]$JobSat<-NA
happy1[(happy1$Income==999999 | happy1$Income==999998 | happy1$Income==0),]$Income<-NA
happy1[(happy1$WorkHrs==999 | happy1$WorkHrs==998 | happy1$WorkHrs==-1),]$WorkHrs<-NA
```

## 2. Producing Models with Different Filled Dataset


```{r }
# Simulate average R-square
record<-rep(NA,200)
record.coeff<-matrix(NA,nrow=200,ncol=13)


# Prepare avaliable data list
columnsneedfixed<-1:12 #Children

#generate a list with all avaiable numbers with the same Happy Value
hsublist1<-list()
hsublist2<-list()
hsublist3<-list()

for (i in 1:3){
  hv<-happy1[happy1$Happy==i,]
  hsublist<-list()
  for (h in columnsneedfixed){
  hsubcol=hv[[h]]
  hsublist[[h]]<-hsubcol[is.na(hsubcol)==FALSE]
  if (i==1){
    hsublist1<-hsublist
  }else if (i==2){
    hsublist2<-hsublist
  }else{
    hsublist3<-hsublist
  }
}
}

#List used to save the Models:
models<-list()


set.seed(7) # For reproductivity
for (s in 1:200){
expandhappy2<-happy1

#Go through each row to fill in the value
for (i in 1:nrow(expandhappy2)){
  # Find the subgroup with the same happy level
  if (expandhappy2[i,13]==1){
    tempsub=hsublist1
  }else if  (expandhappy2[i,13]==2){
    tempsub=hsublist2
  }else{
    tempsub=hsublist3
  }
  
  #Fill in NAs with random valid entry from the corresponding column in the subgroup
  for (j in columnsneedfixed){
    if (is.na(expandhappy2[i,j])){
      expandhappy2[i,j]=sample(tempsub[[j]],size=1)
    }
  }
}

#Data Treatment for Transformation: Add 1 to children
expandhappy2$Children<-expandhappy2$Children+1
#Data Treatment for Transformation: Change 0 Workhrs to 0.1 for transformation
expandhappy2$WorkHrs[expandhappy2$WorkHrs==0]<-0.1
#Data Treatment for Transformation: Change 0 Education to 0.1 for transformation
expandhappy2$Education[expandhappy2$Education==0]<-0.1


# Construct the model
m1<-with(expandhappy2,lm(Happy~Household+Health+OwnHome+Instagram+
                           Marital+Sex+Age+Children+Education+JobSat+Income+WorkHrs))
# Save the summary of model
summ1<-summary(m1)

# Record the R-square and the co-efficients
record[s]<-summ1$adj.r.squared
record.coeff[s,]<-summ1$coefficients[,1]

# Save the model 
models[[s]]<-expandhappy2
}
```

## Demonstration: Distribution of Coefficients

```{r}
# Plot the distributions of all the coefficients from simulation
par(mfrow=c(4,3))
par(mar=c(2,2,2,2))
for (i in 1:12){
  s<-paste("Distribution of estimated beta",as.character(i))
  plot(density(record.coeff[,i]),main=s)
}
```

## Demonstration: Mean Adjusted R sqaure

```{r}
# Average: R^2
meanr<-mean(record)
mean(record)
```

## Demonstration: Distribution of Adjusted R Square

```{r}
plot(density(record))
```

## Model Selection

```{r}
#get the model with r square close to the mean R square
min=1
for (i in 2:200){
  if (abs(record[i]-meanr)<abs(record[min]-meanr)){
    min=i
  }
}
```

## Illustration: Selected Full Model(Model B)

$\hat{Happy}=0.9821-0.04877*Household+0.1842*Health+0.1058*OwnHome+0.07017*Instagram+0.05692*Marital+0.02008*Sex+0.00097*Age+0.1179*Children-0.00437*Education+0.1325*JobSat-0.000002*Income-0.00403*WorkHrs$

```{r}
# Get the dataset for illustration purpose
demo<-models[[min]]
modeldemo<-with(demo,lm(Happy~Household+Health+OwnHome+Instagram+
                          Marital+Sex+Age+Children+Education+
                          JobSat+Income+WorkHrs))

summary(modeldemo)
```

## Plot: Full Model

```{r}
par(mfrow=c(2,2))
plot(modeldemo)
```

## Transformation of Model (Rejected after our analysis)

$\hat{Happy}=0.9821-0.04877*Household^{0.7389}+0.1842*Health^{-0.2865}+0.1058*OwnHome^{-1.6347}+0.07017*Instagram^{3.5317}+0.05692*Marital^{0.3273}+0.02008*Sex+0.00097*Age^{0.3273}+0.1179*log(Children)-0.00437*Education^{1.221}+0.1325*JobSat^{0.2074}-0.000002*Income^{0.1921}-0.00403*WorkHrs$

```{r}
library(alr3)
t_demo_power<-powerTransform(cbind(Happy,Household,Health,OwnHome,Instagram,
                          Marital,Sex,Age,Children,Education,
                          JobSat,Income,WorkHrs)~1,data=demo)
summary(t_demo_power)
t_Happy=demo$Happy^0.7389
t_Household=demo$Household^(-0.2865)
t_Health=demo$Health^0.4839
t_OwnHome=demo$OwnHome^(-1.6347)
t_Instagram=demo$Instagram^3.5317
t_Marital=demo$Marital^0.0654
t_Age=demo$Age^0.3273
t_Children=log(demo$Children)
t_Education=demo$Education^1.221
t_JobSat=demo$JobSat^0.2074
t_Income=demo$Income^0.1921

t_modeldemo<-with(demo,lm(t_Happy~t_Household+t_Health+t_OwnHome+t_Instagram+
                          t_Marital+Sex+t_Age+t_Children+t_Education+
                          t_JobSat+t_Income+WorkHrs))
```

## Summary and Plots of Transformed Model

```{r}
summary(t_modeldemo)
par(mfrow=c(2,2))
plot(t_modeldemo)
```

## Variable Selection Based on Full Model

### Backward Selection Based on AIC

```{r}
backAICdemo <- step(modeldemo,direction="backward", data=demo)
```

### Forward Selection Based on AIC

```{r}
basedemo<- lm(Happy~1,data=demo)
forwardAICdemo<- step(basedemo,scope=list(lower=~1,upper=~Household+Health+OwnHome
                                          +Instagram+Marital+Sex+Age+Children
                                          +Education+JobSat+Income+WorkHrs),
                      direction="forward", data=demo)
```

### Backward Selection Based on BIC

```{r}
backBICdemo <- step(modeldemo,direction="backward", data=demo, k= log(12))
```

### Forward Selection Based on BIC

```{r}
basedemo<- lm(Happy~1,data=demo)
forwardAICdemo<- step(basedemo,scope=list(lower=~1,upper=~Household+Health+OwnHome
                                          +Instagram+Marital+Sex+Age+Children
                                          +Education+JobSat+Income+WorkHrs), 
                      direction="forward", data=demo, k= log(12))
```


## The Final Candidate Model: Model C (Since all selection results are same)


$\hat{Happy}=1.001-0.05437*Household+0.1877*Health+0.1034*OwnHome+0.07328*Instagram+0.05433*Marital+0.0165*Children+0.132*JobSat-0.000002*Income-0.004084*WorkHrs$


```{r}
modelfinal<-with(demo,lm(Happy~Household+Health+OwnHome+Instagram+
                          Marital+Children+JobSat+Income+WorkHrs))
par(mfrow=c(2,2))
plot(modelfinal)
summary(modelfinal)
```

## Partial F-test for the candidate model and full model


```{r}
anova(modelfinal,modeldemo)
```

The p-value shows that our reduced model is significant under a 0.05 significant level.

## Extra: Model A: Based on the dataset where "Don't know" and "Not Answered" are filtered 

$\hat{Happy}=\beta_0-\beta_1*Household+\beta_2*Health+\beta_3*OwnHome+\beta_4*Instagram+\beta_5*Marital+\\x\beta_6*Sex+\beta_7*Age+\beta_8*Children+\beta_9*Education+\beta_{10}*JobSat-\beta_{11}*Income-\beta_{12}*WorkHrs$

```{r}
setwd("~/Downloads/STATS 101A")
happy<-read.table("Happiness.txt",header=T)
library(dplyr)
happy2 <- happy %>% filter(Happy!=8&Happy!=9, Household!=8&Household!=9,
                           Health!=8&Health!=9, OwnHome!=8 &OwnHome!=9, 
                           Instagram!=8 &Instagram!=9, Marital!=9,
                           Age!=98&Age!=99, Children!=9, Education!=98&Education!=99,
                           JobSat!=8&JobSat!=9, Income!=999998&Income!=999999, WorkHrs!=998 &WorkHrs!=999)

modelwhole<-with(happy2,lm(Happy~Household+Health+OwnHome+Instagram+
                     Marital+Sex+Age+Children+Education+JobSat+Income+WorkHrs))
summary(modelwhole)
par(mfrow=c(2,2))
plot(modelwhole)
```
